Vibronic Coupling Effects in the Photoelectron Spectrum of Ozone: A Coupled-Cluster Approach

### 2024-08-18 article.tex
Dr. Stanton's "part"

\begin{abstract}
One of the most important areas of application of equation-of-motion coupled
cluster (EOM-CC) theory is to the prediction, simulation and analysis of
various types of electronic spectra.   In this work, the EOM-CC method known
as EOMIP-CC is applied to the closely lying and coupled pair of states of the
ozone cation -- ${\tilde X}^2A_1$ and ${\tilde A}^2B_2$ -- using sophisticated
approaches that extend to the full singles, doubles, triples and quadruples
model (EOMIP-CCSDTQ).   Combined with a venerable and powerful method for
calculating vibronic spectra from the Hamiltonian produced by EOMIP-CC
calculations, this research provides a spectrum that is in good agreement with
the photoelectron spectrum of ozone.   An important result here is that the
calculations suggest that the adiabatic gap separating these two electronic
states is somewhat smaller than currently thought; an assignment of the
simulated spectrum together with the more precise band positions of the
experimental measurements suggests that this energy gap is 1366$\pm$65
cm$^{-1}$.
\end{abstract}

@@

Amongst the brotherhood of triatomic molecules, it cannot be argued that water
(H$_2$O) is the most important, the most highly studied, and the most well
understood.  Beyond H$_2$O, there are many triatomic molecules that have an
environmental, technological or biological importance, have been subjected to
many studied and are understood to various levels of detail.  Perhaps the most
interesting such case is ozone (O$_3$), which has a vast number of important
properties, a very rich history of study \cite{chappuis}, and – unlike the
relatively simple water molecule – a profound quantum-mechanical complexity
\cite{Babikov:anomalousOzone:2003}.  In the latter context, while we think of
(and an NMR experiment would reveal) as two distinct kinds of oxygen Atoms in
ozone, the full molecular Hamiltonian does not distinguish between the two
types, with the three equivalent structures separated by a barrier that lies
tantalizingly close to the O$_3$ $\rightarrow$ O$_2$ + O dissociation
threshold (102.4 kJ/mol) \cite{Ruscis:ATcT:2022}. In reality, the energy
levels of ozone all have a near triple ($e+a$) degeneracy, albeit with a
tunneling splitting so small that it can be ignored, along with a
semi-infinite lifetime (despite opposing opinion
\cite{Boggs:BerryOzone:2006}).   More than a half century ago, this intriguing
aspect of the ozone molecule was first discussed by Berry
\cite{Berry:Ozone:1960}.

Beyond the structural aspects of ozone, other mysteries center around this
curious molecule.  For example, the distribution of the eighteen distinct
$^{16}$O/$^{17}$O/$^{18}$O isotopologues in the Earth’s atmosphere differs
from what is expected based on the natural isotopic abundance, a puzzle that
has been open for more than three decades
\cite{Mauersberger:OzoneMystery:1990}.

Amongst quantum chemists, ozone has a notorious history, with its strong
biradical character causing significant difficulties in calculations of its
ostensibly simple ground state molecular properties.  An early 1989 study
\cite{Stanton:Ozone:1989, Stanton:Ozone:1989b} by the Bartlett group and
collaborators found that the CCSD+T(CCSD) method predicted that the molecular
equilibrium structure of ozone would have $C_s$ symmetry (that is, the
asymmetric stretching harmonic frequency predicted by this method was
imaginary), a finding that led to a search for better treatment of
non-iterative triple excitations, ultimately leading to the well-known CCSD(T)
treatment \cite{Raghavachari:89, Urban:ccsd(t):1985, GenrefCCSD(T):93}. While
CCSD(T) and higher-level coupled-cluster methods available today
\cite{kucharski:ccsdtq:1992, Kallay:CCHigh, Matthews:ncc:2015} do a good job
in predicting the molecular vibrational potential, an elaborate multireference
configuration interaction study by Dawes {\it et al.} has done an excellent
job on the entire ground-state surface out to the dissociation limit
\cite{Dawes:ozone:2013}. As such, the quantum chemical understanding and
fidelity for the ground electronic state of O$_3$ is now at a mature level.

Qualitatively, the challenge posed to electronic structure by ozone ultimately
arises from its closely spaced highest-occupied ($b_2$) and lowest-unoccupied
($a_1$) molecular orbitals (HOMO and LUMO, respectively).  The two electron
configurations $[\cdots]b_2^2 a_1^0$ and $[\cdots]b_2^0 a_1^2$ mix strongly,
posing the aforementioned challenges with (especially single-reference)
quantum chemical methods.   Another consequence of the identity and energetic
proximity of these two orbitals is that the ozone cation (which is
isoelectronic to the NO$_2$ radical) has closely lying $^2A_1$ and $^2B_2$
electronic states.   Like the associated states in NO$_2$, both of these
states are plagued by orbital symmetry-breaking problems
\cite{Davidson:SymmBreak:76}, a problem that greatly complicates
quantum-chemical efforts to them.   One of the many accomplishments of the
Bartlett group has been the integral role played by them in the development of
equation-of-motion coupled-cluster theory \cite{Stanton:93:EOMCC,
Nooijen:EOMEA:95, Bartlett:CC_review:07} (EOM-CC, also known as linear
response coupled-cluster theory \cite{Koch:90:LinResp}). These methods provide
a very efficient and simple way to study certain classes of what are often
termed ``multireference problems" \cite{Krylov:EOMRev:07}, and are ideally
suited to studying many reactive intermediates, radicals, biradicals and
electronically excited states.   From a somewhat wider viewpoint, the
existence of closely spaced electronic states always carries the potential for
(possibly strong) vibronic coupling, a phenomenon that can play an important
role in molecular dynamics and spectroscopy.  Indeed, one of the great
successes of EOM-CC methods has been in their ability -- if and only if
combined with vibronic coupling models -- to enable high-quality simulations
of complicated electronic spectra.   Such work provides important insight into
the nature of vibronic coupling in molecular systems, as has been exemplified
by various application studies (for example, see Refs.~\cite{Stanton:NO3:07}
and~\cite{Koppel:02}).

Our contribution to this issue paying homage to the career and accomplishments
of R.J. Bartlett consists of an application of a vibronic coupling model
parametrized by EOM-CC calculations to the photoelectron spectrum of ozone.
We trust that this combination of methodology applied to a molecule that has
been extensively studied by the Bartlett group is an appropriate contribution
to this issue.

>>>>

\begin{abstract}
One of the most important areas of application for equation-of-motion coupled
cluster (EOM-CC) theory is to the prediction, simulation and analysis of
various types of electronic spectra.   In this work, the EOM-CC method known
as EOMIP-CC is applied to the closely lying and coupled pair of states of the
ozone cation -- ${\tilde X}^2A_1$ and ${\tilde A}^2B_2$ -- using sophisticated
approaches that extend to the full singles, doubles, triples and quadruples
model (EOMIP-CCSDTQ).   Combined with a venerable and powerful method for
calculating vibronic spectra from the Hamiltonian produced by EOMIP-CC
calculations, this research provides a spectrum that is in good agreement with
the photoelectron spectrum of ozone.   An important result here is that the
calculations suggest that the adiabatic gap separating these two electronic
states is somewhat smaller than currently thought; an assignment of the
simulated spectrum together with the more precise band positions of the
experimental measurements suggests that this energy gap is 1366$\pm$65
cm$^{-1}$.
\end{abstract}

@@

Amongst the brotherhood of triatomic molecules, it cannot be denied that water
(H$_2$O) is the most important, the most highly studied, and the most
well-understood.  Beyond H$_2$O, there are many triatomic molecules that have
an environmental, technological or biological importance, have been subjected
to many studied and are understood to various levels of detail.  Perhaps the
most interesting such case is ozone (O$_3$), which has a vast number of
important properties, a very rich history of study \cite{chappuis}, and –
unlike the relatively simple water molecule – a profound quantum-mechanical
complexity \cite{Babikov:anomalousOzone:2003}.  In the latter context, while
we think of (and an NMR experiment would reveal) as two distinct kinds of
oxygen atoms in ozone, the full molecular Hamiltonian does not distinguish
between them types, with the three equivalent structures separated by a
barrier that lies tantalizingly close to the O$_3$ $\rightarrow$ O$_2$ + O
dissociation threshold (102.4 kJ/mol) \cite{Ruscis:ATcT:2022}. In reality, the
energy levels of ozone all have a near triple ($e+a$) degeneracy, albeit with
a tunneling splitting so small that it can be ignored, along with a
semi-infinite lifetime (despite opposing opinion
\cite{Boggs:BerryOzone:2006}).   More than a half century ago, this intriguing
aspect of the ozone molecule was first discussed by Berry
\cite{Berry:Ozone:1960}.

Beyond the structural aspects of ozone, other mysteries center around this
curious molecule.  For example, the distribution of the eighteen distinct
$^{16}$O/$^{17}$O/$^{18}$O isotopologues in the Earth's atmosphere differs
from what is expected based on the natural isotopic abundance, a puzzle that
has been open for more than three decades
\cite{Mauersberger:OzoneMystery:1990}.

Among quantum chemists, ozone has a notorious history, with its strong
biradical character causing significant difficulties in calculations of its
ostensibly simple ground state molecular properties.  An early 1989 study
\cite{Stanton:Ozone:1989, Stanton:Ozone:1989b} by the Bartlett group and
collaborators found that the CCSD+T(CCSD) method predicted that the molecular
equilibrium structure of ozone would have $C_s$ symmetry (that is, the
asymmetric stretching harmonic frequency predicted by this method is
imaginary), a finding that led to a search for better treatment of
non-iterative triple excitations, ultimately leading to the well-known CCSD(T)
treatment \cite{Raghavachari:89, Urban:ccsd(t):1985, GenrefCCSD(T):93}. While
CCSD(T) and higher-level coupled-cluster methods available today
\cite{kucharski:ccsdtq:1992, Kallay:CCHigh, Matthews:ncc:2015} do a good job
in predicting the molecular vibrational potential, an elaborate multireference
configuration interaction study by Dawes {\it et al.} has done an excellent
job on the entire ground-state surface out to the dissociation limit
\cite{Dawes:ozone:2013}. As such, the quantum chemical understanding and
fidelity for the ground electronic state of O$_3$ is now at a mature level.

Qualitatively, the challenge posed to electronic structure theory by ozone
ultimately arises from its closely spaced highest-occupied ($a_2$) and
lowest-unoccupied ($b_1$) molecular orbitals (HOMO and LUMO, respectively).
The two electron configurations $[\cdots]b_2^2 a_1^0$ and $[\cdots]b_2^0
a_1^2$ mix strongly, posing the aforementioned challenges with (especially
single-reference) quantum chemical methods.   A second feature consequence of
the identity and energetic proximity of these two orbitals is that the ozone
cation (which is isoelectronic to the NO$_2$ radical) has closely lying
$^2A_1$ and $^2B_2$ electronic states.   Like the associated states in NO$_2$,
both of these states are plagued by orbital symmetry-breaking problems
\cite{Davidson:SymmBreak:76}, a problem that greatly complicates
quantum-chemical efforts to characterize them.   One of the many
accomplishments of the Bartlett group has been the integral role played by
them in the development of equation-of-motion coupled-cluster theory
\cite{Stanton:93:EOMCC, Nooijen:EOMEA:95, Bartlett:CC_review:07} (EOM-CC, also
known as linear response coupled-cluster theory \cite{Koch:90:LinResp}). These
methods provide a very efficient and simple way to study certain classes of
what are often termed ``multireference problems" \cite{Krylov:EOMRev:07}, and
are ideally suited to studying many reactive intermediates, radicals,
biradicals and electronically excited states.   From a somewhat wider
viewpoint, the existence of closely spaced electronic states always carries
the potential for (possibly strong) vibronic coupling, a phenomenon that can
play an important role in molecular dynamics and spectroscopy.  Indeed, one of
the great successes of EOM-CC methods has been in their ability -- if and only
if combined with vibronic coupling models -- to enable high-quality
simulations of complicated electronic spectra.   Such work provides important
insight into the nature of vibronic coupling in molecular systems, as has been
exemplified by various application studies (for example, see
Refs.~\cite{Stanton:NO3:07} and~\cite{Koppel:02}).

Our contribution to this issue paying homage to the career and accomplishments
of R.J. Bartlett consists of an application of a vibronic coupling model,
parametrized by EOM-CC calculations, to the photoelectron spectrum of ozone.
We trust that this combination of methodology applied to a molecule that has
been extensively studied by the Bartlett group is an appropriate contribution
to this issue.
